---
title: ""
output: 
  pdf_document: 
    keep_tex: yes
---

```{r setup, results='hide', message=FALSE, warning=FALSE}
library(haven)
library(MatchIt)
library(stargazer)
library(tidyverse)

df <- read_dta("lalonde2.dta")

#################
### Functions ###
#################

# Loop over ttests
testing <- function(data, var){
  temp0 = filter(data, variables==var & treated==0) %>% 
    select(values)
  temp1 = filter(data, variables==var & treated==1) %>% 
    select(values)
  test = t.test(temp0, temp1)
  output <- data.frame(var, round(test$estimate[1], 3), 
                       round(test$estimate[2], 3), round(test$p.value, 3))
  return(output)
}

# Create t.test table
test.table <- function(data, vars, filename) {
  tests <- lapply(vars, testing, data=data)
  tests <- data.table::rbindlist(tests)
  table <- knitr::kable(tests, "latex", vline="", linesep="", 
                        toprule="\\hline \\hline", bottomrule = "\\hline \\hline", 
                        col.names=c("Variable", "Control", 
                                    "Treated", "$p$-value"), 
                        escape=F)
  writeLines(table, filename)
}

# Plot covariate balance
covar.balance <- function(data, vars, color, filename){
  df <- data %>% 
  group_by(treated, variables) %>% 
  summarize(avg = mean(values),
            std = sd(values),
            num = n()) %>% 
  mutate(se = std/sqrt(num)) %>% 
  filter(variables %in% vars)
  
  p <- ggplot(df, aes(x=treated, group=treated)) + 
    geom_point(aes(y=avg), color=color) + 
    geom_errorbar(aes(ymin=avg-1.96*se, ymax=avg+1.96*se), width=.2,
                 position=position_dodge(.9), color=color) +
    facet_wrap( ~ variables, scales="free", ncol=3) + 
    theme_minimal() +
    labs(x="Treated", y="Values")
  ggsave(p, filename=filename)
}

# Smooth out earnings by propensity score using local linear regression
loclin <- function(data, t){
  data <- filter(data, treated == t)
  lp <- loess(re78 ~ prop_score, data = data, degree = 1)
  data$smoothed_re78 <- predict(lp)
  return(data)
}

```

What covariates do we care about?
```{r,results='hide', message=FALSE, warning=FALSE}
# Specify covariates
covars <- c("age", "black", "hisp", "educ", "married", "re74", "kids18", 
            "kidmiss", "nodegree", "w76", "re75")
newnames <- c("Age", "Black", "Hispanic", "Education", "Married", 
              "Earnings in 1974", "Kids 18", "Kid Miss", "No Degree",
              "Employed in 1976", "Earnings in 1975")
cont_covars <- c("Age", "Education", "Earnings in 1974", "Kids 18", 
                 "Earnings in 1975")
```

Where do we cut off propensity scores?
```{r,results='hide', message=FALSE, warning=FALSE}
phigh=0.95
plow=0.05

phigh_alt = 0.6
plow_alt = 0.1

final_high = phigh_alt
final_low = plow

```


Pretty colors:
```{r,results='hide', message=FALSE, warning=FALSE}
color.pal <- c("#CADF9E", "#001B2E", "#294C60", "#087E8B", "#7F636E")
```


## A: Test covariate balance in NSW sample
```{r,results='hide', message=FALSE, warning=FALSE}
df_experimental <- df %>% 
  filter(sample==1)

df_exp_long <- df_experimental %>% 
  pivot_longer(-c("treated", "idnum", "sample"), names_to = "variables", 
               "values_to" = "values") %>% 
  mutate(treated=as.factor(treated)) %>% 
  mutate(variables=plyr::mapvalues(variables, from=covars, to=newnames))

# test table
test.table(df_exp_long, newnames, filename="ps2_q2a.tex")

# plots
covar.balance(df_exp_long, newnames, color.pal[3], "ps2_q2a.pdf")
```

## B: Estimating effect using experimental sample

```{r,results='hide', message=FALSE, warning=FALSE}
att1 <- lm(re78 ~ treated, data=df_experimental)
control_formula <- paste("re78 ~ treated +", paste(covars, collapse="+"))
att2 <- lm(control_formula, data=df_experimental)
stargazer(att1, att2, covariate.labels=c("ATT", "Constant"), no.space=T,
          header=F, keep.stat="n", dep.var.caption="", float=F,
          dep.var.labels.include=F, omit=covars, 
          star.cutoffs = c(0.1, 0.05, 0.01), 
          add.lines = list(c("Controls", "No", "Yes")),
          out="ps2_q2b.tex")
```


## C: Estimate effect with NSW and CPS samples

CPS as controls:
```{r,results='hide', message=FALSE, warning=FALSE}
df_cps <- df %>% 
  filter(sample<= 2 & (treated==1 | is.na(treated))) %>% 
  mutate(treated = ifelse(is.na(treated), 0, treated))

att_cps1 <- lm(re78 ~ treated, data=df_cps)

control_formula <- paste("re78 ~ treated +", paste(covars, collapse="+"))
att_cps2 <- lm(control_formula, data=df_cps)
stargazer(att_cps1, att_cps2, covariate.labels=c("ATT", "Constant"), no.space=T,
          header=F, keep.stat="n", dep.var.caption="", float=F,
          dep.var.labels.include=F, omit=covars, 
          star.cutoffs = c(0.1, 0.05, 0.01), 
          add.lines = list(c("Controls", "No", "Yes")),
          out="ps2_q2c.tex")
```

## D: Investigate covariate balancing and support
```{r,results='hide', message=FALSE, warning=FALSE}
# check for covariate balancing
df_cps_long <- df_cps %>% 
  pivot_longer(-c("treated", "idnum", "sample"), names_to = "variables", 
               "values_to" = "values") %>% 
  mutate(treated=as.factor(treated),
         sample_factor= ifelse(sample==1, "Experimental", "CPS")) %>% 
  group_by(treated, variables) %>% 
  ungroup() %>% 
  mutate(variables=plyr::mapvalues(variables, from=covars, to=newnames))

# test table
test.table(df_cps_long, newnames, filename="ps2_q2d.tex")

# plots
covar.balance(df_cps_long, newnames, color.pal[3], "ps2_q2d_1.pdf")

# check for common support of only continuous variables
df_cps_long <- df_cps_long %>% 
  filter(variables %in% cont_covars)

p <- ggplot(df_cps_long) + 
  geom_histogram(aes(values, after_stat(density), fill=sample_factor), alpha=0.5) +
    scale_fill_manual(values=color.pal[c(1,3)]) +
  facet_wrap( ~ variables, scales="free", ncol=3) + 
  theme_minimal() + theme(legend.position="bottom") +
  labs(x="Values", y="Density", fill="Sample")
ggsave(p, filename="ps2_q2d_2.pdf")
```

## E: 1 nearest neighbor propensity score matching

Step 1: Let's first estimate the propensity score
```{r,results='hide', message=FALSE, warning=FALSE}
ps_formula <- as.formula(paste("treated ~ ", paste(covars, collapse="+")))
# draw with replacement
ps_match <- matchit(ps_formula, data = df_cps,
                 method = "nearest", distance = "glm", replace=T, link="probit")
df_cps <- cbind(df_cps, prop_score=ps_match$distance)
```

Step 2: Now check common support
```{r,results='hide', message=FALSE, warning=FALSE}
p <- ggplot() + 
  geom_histogram(data=filter(df_cps, treated==1), 
                 aes(x = prop_score, y=after_stat(density)),
                 fill=color.pal[1], alpha=0.7) + 
  geom_histogram(data=filter(df_cps, treated==0), 
                 aes(x = prop_score, y=-after_stat(density)),
                 fill=color.pal[3], alpha=0.7) +
  theme_minimal() +
  labs(x="Values", y="Density")
ggsave(p, filename="ps2_q2e_1.pdf")

# How does it look trimmed?
p <- ggplot() + 
  geom_histogram(data=filter(df_cps, treated==1 & prop_score>plow & 
                               prop_score<phigh), 
                 aes(x = prop_score, y=after_stat(density)),
                 fill=color.pal[1], alpha=0.7) + 
  geom_histogram(data=filter(df_cps, treated==0 & prop_score>plow & 
                               prop_score<phigh), 
                 aes(x = prop_score, y=-after_stat(density)),
                 fill=color.pal[3], alpha=0.7) +
  theme_minimal() +
  labs(x="Values", y="Density")
ggsave(p, filename="ps2_q2e_2.pdf")

# Alternate trimming:
p <- ggplot() + 
  geom_histogram(data=filter(df_cps, treated==1 & prop_score>plow_alt & 
                               prop_score<phigh_alt), 
                 aes(x = prop_score, y=after_stat(density)),
                 fill=color.pal[1], alpha=0.7) + 
  geom_histogram(data=filter(df_cps, treated==0 & prop_score>plow_alt & 
                               prop_score<phigh_alt), 
                 aes(x = prop_score, y=-after_stat(density)),
                 fill=color.pal[3], alpha=0.7) +
  theme_minimal() +
  labs(x="Values", y="Density")
ggsave(p, filename="ps2_q2e_3.pdf")
```

Step 3: One to one matching with replacement 
```{r,results='hide', message=FALSE, warning=FALSE}
df_knn <- match.data(ps_match) %>% 
  filter(prop_score>final_low & prop_score<final_high)
p <- ggplot() + 
  geom_histogram(data=df_knn, 
                 aes(x = prop_score, y=after_stat(density)),
                 fill=color.pal[1], alpha=0.7) + 
  geom_histogram(data=df_knn, 
                 aes(x = prop_score, y=-after_stat(density)),
                 fill=color.pal[3], alpha=0.7) +
  theme_minimal() +
  labs(x="Values", y="Density")
ggsave(p, filename="ps2_q2e_4.pdf")

# ensure covariate balance
df_knn_long <- df_knn %>% 
  pivot_longer(-c("treated", "idnum", "sample"), names_to = "variables", 
               "values_to" = "values") %>% 
  mutate(treated=as.factor(treated),
         sample_factor= ifelse(sample==1, "Experimental", "CPS")) %>% 
  group_by(treated, variables) %>% 
  mutate(variables=plyr::mapvalues(variables, from=covars, to=newnames))

covar.balance(df_knn_long, newnames, color.pal[3], "ps2_q2e_5.pdf")
```

Step 4: Estimate ATT
```{r,results='hide', message=FALSE, warning=FALSE}
ps_att1 <- lm(re78 ~ treated, data=df_knn, weights=weights)

control_formula <- paste("re78 ~ treated +", paste(covars, collapse="+"))

ps_att2 <- lm(control_formula, data=df_knn, weights=weights)
stargazer(ps_att1, ps_att2, covariate.labels=c("ATT", "Constant"), no.space=T,
          header=F, keep.stat="n", dep.var.caption="", float=F,
          dep.var.labels.include=F, omit=covars, 
          star.cutoffs = c(0.1, 0.05, 0.01), 
          add.lines = list(c("Controls", "No", "Yes")),
          out="ps2_q2e.tex")
```


## F: Local linear regression

```{r,results='hide', message=FALSE, warning=FALSE}
df_loclin <- bind_rows(loclin(df_knn, 1), loclin(df_knn, 0)) 

# ATT
ll_att1 <- lm(smoothed_re78 ~ treated, data=df_loclin)
ll_att2 <- lm(smoothed_re78 ~ treated, data=df_loclin, weights=weights)

control_formula <- paste("re78 ~ treated +", paste(covars, collapse="+"))

ll_att3 <- lm(control_formula, data=df_knn)
stargazer(ll_att1, ll_att2, ll_att3, covariate.labels=c("ATT", "Constant"), no.space=T,
          header=F, keep.stat="n", dep.var.caption="", float=F,
          dep.var.labels.include=F, omit=covars, 
          star.cutoffs = c(0.1, 0.05, 0.01), 
          add.lines = list(c("Controls", "No", "No", "Yes"),
                           c("Propensity Weights", "No", "Yes", "No")),
          out="ps2_q2f.tex")

# Plot predicted earnings
df_loclin_long <- df_loclin %>% 
  select(treated, smoothed_re78, re78, prop_score) %>% 
  pivot_longer(-c("treated", "prop_score"), names_to = "variables", 
               "values_to" = "values") %>% 
  mutate(treated=as.factor(treated),
         variables=ifelse(variables=="re78", "Raw Earnings", "Smoothed Earnings") %>% as.factor) 

p <- ggplot(data=filter(df_loclin_long, treated==1)) + 
  geom_point(aes(x=prop_score, y=values, color=variables)) +
  scale_color_manual(values=color.pal[c(3,5)]) +
  theme_minimal() + theme(legend.position="bottom") +
  labs(x="Earnings", y="Propensity Score", color="Variable")
ggsave(p, filename="ps2_q2f_1.pdf")

p <- ggplot(data=filter(df_loclin_long, treated==0)) + 
  geom_point(aes(x=prop_score, y=values, color=variables)) +
  scale_color_manual(values=color.pal[c(3,5)]) +
  theme_minimal() + theme(legend.position="bottom") +
  labs(x="Earnings", y="Propensity Score", color="Variable")
ggsave(p, filename="ps2_q2f_2.pdf")

```

